Faster generic IND-CCA2 secure KEM using ``encrypt-then-MAC''

### 2024-10-03 etm-transform/manuscript/main.tex
Drafted introduction

%%%% 2. PACKAGES %%%%
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage{multirow}
% \usepackage[demo]{graphicx}

@@

%%%% 7. PAPER CONTENT %%%%
\section{Introduction}
Key encapsulation mechanism (KEM) is a cryptographic primitive that allows two parties to establish a shared secret over an insecure channel. The combination of KEM and some data encapsulation mechanism (DEM), such as AES-GCM and ChaCha20-Poly1305, is the foundation of many of today's most popular secure communication protocols such as Transport Layer Security (TLS) and Secure Shell (SSH). The commonly accepted security standard of a KEM is \emph{Indistinguishability under adaptive chosen-ciphertext attack (IND-CCA2)}: no efficient adversary can distinguish a shared secret obtained by running the encapsulation routine from random noise, even with access to a decapsulation oracle throughout the attack. KEM is related to another important cryptographic primitive called public-key encryption (PKE), and their security standards are similar. The desired security standard for PKE, also called IND-CCA2, requires that no efficient adversary can distinguish the encryption of two adversarially chosen messages even with access to decryption oracle.

It is difficult to build a provably IND-CCA2 secure KEM from scratch. Instead, secure KEMs are usually built on top of a PKE with weaker security property (e.g. being only OW-CPA or IND-CPA secure, not IND-CCA2). One such construction is the Fujisaki-Okamoto transformation, proposed in 1999 by Fujisaki Eiichiro and Okamoto Tetsuyaki in their seminal paper \cite{DBLP:conf/crypto/FujisakiO99}\cite{DBLP:journals/joc/FujisakiO13}. The first Fujisaki-Okamoto transformation combines an OW-CPA secure PKE with an IND-CPA secure symmetric cipher into a hybrid PKE (HPKE) with proven IND-CCA2 security under the random oracle model. Subsequent proposals such as \cite{DBLP:conf/ctrsa/OkamotoP01}, \cite{DBLP:conf/ctrsa/CoronHJPPT02}, and \cite{DBLP:conf/ima/Dent03} improved on the original proposal and adapted it to build KEM instead of HPKE. This line of work culminated in a landmark publication in 2017 by Hofhein, Hovelmann, and Kiltz \cite{DBLP:conf/tcc/HofheinzHK17}\cite{DBLP:conf/asiacrypt/HovelmannsHM22}, where the authors provided a versatile variety of modular KEM constructions with tight security reduction in the random oracle model and non-tight security reduction in the quantum random oracle model.

The modular Fujisaki-Okamoto KEM transformation is remarkably successful. It was adopted by many submissions to NIST's post-quantum cryptography competition, including Kyber \cite{DBLP:conf/eurosp/BosDKLLSSSS18}, Saber \cite{DBLP:conf/africacrypt/DAnversKRV18}, FrodoKEM \cite{DBLP:conf/ccs/BosCDMNNRS16}, and classic McEliece \cite{classicmceliecespec} among others. When Kyber was standardized by NIST in FIPS 203 ``Module-lattice key-encapsulation mechanism'' (ML-KEM) \cite{FIPS203}, it kept the Fujisaki-Okamoto transformation in its KEM construction. However, the Fujisaki-Okamoto transformation is not perfect. Among its shortcomings is the use of \emph{de-randomization} (transform a randomized encryption routine into a deterministic routine by using a pseudorandom coin derived from the input message) and \emph{re-encryption} (the decapsulation routine of the output KEM runs the encryption routine of the input PKE to ensure ciphertext integrity), which causes two problems: \begin{itemize}
    \item \textbf{computational inefficiency:} where the PKE's encryption routine is substantially more expensive than the decryption routine, using re-encryption causes the decapsulation routine in the output KEM to become computationally expensive
    \item \textbf{side-channel vulnerability:} running the input PKE's encryption routine in the output KEM's decapsulation routine introduces risk of side-channel vulnerabilities not found in the input PKE's decryption routine alone. In fact, many practical attacks \cite{ueno2022curse}\cite{ravi2019generic} exploit re-encryption to decrypt ciphertext or recover secret keys. Countermeasures such as masking have been proposed to address these side channels, but they inevitably carry substantial performance penalty.
\end{itemize}

\subsection{Our contributions}
Our main contribution is a novel generic construction that combines a PKE with weaker security property and a one-time secure message authentication code (MAC) into an IND-CCA2 secure KEM in the random oracle model. Specifically, we require the input PKE to be one-way secure against plaintext-checking attacks (OW-PCA): no efficient adversary can recover the decryption of a random encryption with access to a plaintext-checking oracle (PCO) that, when queried on a plaintext-ciphertext pair $(m, c)$, answers $m$ is the decryption of $c$. While OW-PCA is not a standard security definition, it has appeared useful in past IND-CCA2 KEM constructions \cite{DBLP:conf/ctrsa/OkamotoP01}\cite{DBLP:conf/ctrsa/CoronHJPPT02}, and plays an important role in the modularity of the Fujisaki-Okamoto KEM transformation.

Our KEM construction is inspired by how symmetric cryptography achieves IND-CCA2 security (or equivalently, authenticated encryption \cite{DBLP:conf/crypto/Krawczyk01}) using the ``encrypt-then-MAC'' pattern. We applied ``encrypt-then-MAC'' to the construction of the KEM: at encapsulation, a symmetric key is derived from hashing a random PKE plaintext, then used to compute an authenticator against the PKE ciphertext; at decapsulation, the same symmetric key is derived from hashing the decryption of the PKE ciphertext, then used to verify the authenticator. Intuitively, for an adversary to be able to produce a valid authenticator, it must know the correct symmetric key and thus the corresponding PKE plaintext. In other words, the adversary either produced the ciphertext honestly, or have broken the one-wayness of the input PKE.

\subsubsection{Performance improvements}
The main advantage of our KEM construction over the Fujisaki-Okamoto transformation is the performance gains: our construction replaces re-encryption with computing a symmetric message authenticator, which is significantly faster, especially for one-time MAC. At the cost of computing an authenticator in encapsulation and increasing the ciphertext size by a message authenticator, our construction speeds up decapsulation by 10x.

When instantiated with the underlying PKE routines of ML-KEM and the Poly1305 message authenticator, our construction $\text{ML-KEM}^+$ achieves on average 72\%-80\% reduction of CPU cycles needed for decapsulation while only incurring 2\%-7\% increase of CPU cycles for encapsulation when compared to ML-KEM.

We also implemented and measured the round trip time of key exchange protocols with various modes of authentication. When compared to ML-KEM, $\text{ML-KEM}^+$ achieves 24\%-28\% reduction of round trip time in unauthenticated key exchange, 29\%-35\% reduction in unilaterally authenticated key exchange, and 35\%-48\% reduction in mutually authenticated key exchange.

\subsection{Related works}
A similar construction was proposed by Abdulla et al \cite{DBLP:conf/ctrsa/AbdallaBR01}, though our construction builds a KEM instead of a HPKE, and our construction is generic over all OW-PCA secure PKE.

@@

\subsection{Public-key encryption scheme}
\textbf{\textit{Syntax.}} A public-key encryption scheme $\pke(\keygen, \encrypt, \decrypt)$ is a collection of three routines defined over some plaintext space $\mathcal{M}$ and some ciphertext space $\mathcal{C}$. $(\pk, \sk) \leftsample \keygen()$ is a randomized routine that returns a keypair. The encryption routine $\encrypt: (\pk, m) \mapsto c$ encrypts the input plaintext under the input public key. The decryption routine $\decrypt: (\sk, c) \mapsto m$ decrypts the input ciphertext under the input secret key. Where the encryption routine is randomized, we denote the randomness by $r \in \mathcal{R}$, where $\mathcal{R}$ is called the coin space. The decryption routine is assumed to always be deterministic. Some decryption routines can detect malformed ciphertext and output the rejection symbol $\bot$ accordingly.

\textbf{\textit{Correctness.}} Following the definition in \cite{DBLP:conf/eurocrypt/DworkNR04} and \cite{hofheinz2017modular}, a $\pke$ is $\delta$-correct if:

@@

\emph{Security.} We discuss the security of a $\pke$ using the sequence of games described in \cite{shoup2004sequences}. Specifically, we first define the $\texttt{OW-ATK}$ as they pertain to a public key encryption scheme. In later section we will define the $\texttt{IND-CCA}$ game as it pertains to a key encapsulation mechanism.

>>>>

%%%% 2. PACKAGES %%%%
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{positioning}
% \usepackage[demo]{graphicx}

@@

%%%% 7. PAPER CONTENT %%%%
\section{Introduction}\label{sec:introduction}
Key encapsulation mechanism (KEM) is a cryptographic primitive that allows two parties to establish a shared secret over an insecure channel. The combination of KEM and some data encapsulation mechanism (DEM), such as AES-GCM and ChaCha20-Poly1305, is the foundation of many of today's most popular secure communication protocols such as Transport Layer Security (TLS) and Secure Shell (SSH). The commonly accepted security standard of a KEM is \emph{Indistinguishability under adaptive chosen-ciphertext attack (IND-CCA2)}: no efficient adversary can distinguish a shared secret obtained by running the encapsulation routine from random noise, even with access to a decapsulation oracle throughout the attack. KEM is related to another important cryptographic primitive called public-key encryption (PKE), and their security standards are similar. The desired security standard for PKE, also called IND-CCA2, requires that no efficient adversary can distinguish the encryption of two adversarially chosen messages even with access to decryption oracle.

It is difficult to build a provably IND-CCA2 secure KEM from scratch. Instead, secure KEMs are usually built on top of a PKE with weaker security property (e.g. being only OW-CPA or IND-CPA secure, not IND-CCA2). One such construction is the Fujisaki-Okamoto transformation, proposed in 1999 by Fujisaki Eiichiro and Okamoto Tetsuyaki in their seminal paper \cite{DBLP:conf/crypto/FujisakiO99}\cite{DBLP:journals/joc/FujisakiO13}. The first Fujisaki-Okamoto transformation combines an OW-CPA secure PKE with an IND-CPA secure symmetric cipher into a hybrid PKE (HPKE) with proven IND-CCA2 security under the random oracle model. Subsequent works \cite{DBLP:conf/ctrsa/OkamotoP01}\cite{DBLP:conf/ctrsa/CoronHJPPT02}\cite{DBLP:conf/ima/Dent03} improved on the original proposal and adapted it to build KEM instead of HPKE. This line of work culminated in a landmark publication in 2017 by Hofhein, Hovelmann, and Kiltz \cite{DBLP:conf/tcc/HofheinzHK17}\cite{DBLP:conf/asiacrypt/HovelmannsHM22}, where the authors provided a versatile variety of modular KEM constructions with tight security reduction in the random oracle model and non-tight security reduction in the quantum random oracle model.

The modular Fujisaki-Okamoto KEM transformation is remarkably successful. It was adopted by many submissions to NIST's post-quantum cryptography competition, including Kyber \cite{DBLP:conf/eurosp/BosDKLLSSSS18}, Saber \cite{DBLP:conf/africacrypt/DAnversKRV18}, FrodoKEM \cite{DBLP:conf/ccs/BosCDMNNRS16}, and classic McEliece \cite{classicmceliecespec} among others. When Kyber was standardized by NIST in FIPS 203 ``Module-lattice key-encapsulation mechanism'' (ML-KEM) \cite{FIPS203}, it kept the Fujisaki-Okamoto transformation in its KEM construction. However, the Fujisaki-Okamoto transformation is not perfect. As pointed out by the authors of \cite{hofheinz2017modular}, the Fujisaki-Okamoto transformation uses \emph{de-randomization} and \emph{re-encryption} to ensure ciphertext integrity, which brings two problems: \begin{itemize}
    \item \textbf{computational inefficiency:} where the PKE's encryption routine is substantially more expensive than the decryption routine, using re-encryption causes the decapsulation routine in the output KEM to become computationally expensive
    \item \textbf{side-channel vulnerability:} running the input PKE's encryption routine in the output KEM's decapsulation routine introduces risk of side-channel vulnerabilities not found in the input PKE's decryption routine alone. In fact, many practical attacks \cite{ueno2022curse}\cite{ravi2019generic} exploit re-encryption to decrypt ciphertext or recover secret keys. Countermeasures such as masking have been proposed to address these side channels, but they inevitably carry substantial performance penalty.
\end{itemize}

\subsection{Our contributions}
Our main contribution is a novel generic KEM construction that combines a PKE with weak security and a message authentication code (MAC). We were inspired by the strategy used in symmetric cyrptography for achieving IND-CCA2 security: combining a symmetric cipher with a MAC in a pattern called ``encrypt-then-MAC''. In the encapsulation routine of the ``encrypt-then-MAC'' KEM construction, a random PKE plaintext $m$ is sampled and encrypted into some PKE ciphertext $c^\prime$; a symmetric key $k$, derived from hashing the plaintext, is used to compute a message authenticator $t$ on the PKE ciphertext $c^\prime$. The KEM ciphertext $c = (c^\prime, t)$ contains both the PKE ciphertext and the message authenticator. At decapsulation, the PKE ciphertext $c^\prime$ is decrypted into $\hat{m}$ first, then a symmetric key $\hat{k}$ is derived from hashing the decryption $\hat{m}$. The symmetric key is used to verify the authenticator against the PKE ciphertext. A illustration of the data flow in the ``encrypt-then-MAC'' KEM construction can be found in figure \ref{fig:etm-diagram}. A detailed description of the individual routines can be found in figure \ref{fig:etm-routines}.

\begin{figure}
    \centering
    \begin{tikzpicture}
        \tikzstyle{rect} = [draw, rectangle, minimum width=3cm, minimum height=1cm]
        \tikzstyle{filledrect} = [draw, rectangle, minimum width=3cm, minimum height=1cm, fill=lightgray]
        \matrix [column sep=7mm, row sep=5mm] {
            \node (pk) [rect] {$\pk$}; &
            \node (m) [rect] {$m \leftsample \mathcal{M}$}; &
            \node (hashg) [rect] {Hash function $G$}; \\
            \node (pkeenc) [rect] {$\texttt{PKE.Enc}$}; &
            \node (mackey) [rect] {$k_\mac$}; \\
            \node (ct) [filledrect] {$c^\prime$}; &
            \node (mac) [rect] {$\mac$}; &
            \node (tag) [filledrect] {$t$}; \\
        };
        \draw[->, thick] (pk) -- (pkeenc);
        \draw[->, thick] (m) -- (pkeenc);
        \draw[->, thick] (m) -- (hashg);
        \draw[->, thick] (hashg) -- (mackey);
        \draw[->, thick] (pkeenc) -- (ct);
        \draw[->, thick] (mackey) -- (mac);
        \draw[->, thick] (ct) -- (mac);
        \draw[->, thick] (mac) -- (tag);
    \end{tikzpicture}

    \caption{Combining PKE with MAC using ``encrypt-then-MAC'' to encapsulate a shared secret. 
    The returned values are colored grey
    }\label{fig:etm-diagram}
\end{figure}

\subsubsection{Security reduction}
By authenticating the PKE ciphertext using a symmetric key derived from its decryption, the ``encrypt-then-MAC'' construction greatly limits the amount of information an adversary can learn from a chosen-ciphertext attack. For an adversary to produce a valid KEM ciphertext, it must produce a valid message authenticator for the PKE ciphertext under the correct symmetric key. Under the random oracle model, the symmetric key is indistinguishable from uniformly random unless the adversary also knows the corresponding plaintext completely, thus rendering the output of the decasulation oracle redundant. Furthermore, because each encapsulation will pick a random message, within a reasonable lifetime of the KEM keypair, each KEM ciphertext will be authenticated using a distinct symmetric key, so the choice of MAC is only required to be one-time secure, affording us additional performance gains by choosing more efficient MAC instances. In section \ref{sec:main-results}, we prove that the IND-CCA2 security of the ``encrypt-then-MAC'' KEM reduces tightly to the OW-PCA security of the input PKE and the one-time existential unforgeability of the input MAC under the random oracle model.

One-wayness under plaintext-checking attack (OW-PCA)\cite{DBLP:conf/ctrsa/OkamotoP01} requires that no efficient adversary can recover some random encryption even with access to a plaintext-checking oracle, who accepts a plaintext-ciphertext pair $(m, c)$ and returns whether the queried ciphertext decrypts to the queried plaintext. While it is not a standard security notion, it is useful when reducing the security of the ``encrypt-then-MAC'' KEM to the security of the PKE, and there are many well studied cryptosystems that are known to be OW-PCA: the RSA cryptosystem \cite{DBLP:journals/cacm/RivestSA78} is known to be a trapdoor permutation and thus trivially OW-PCA secure; the ElGamal cryptosystem  \cite{DBLP:journals/tit/Elgamal85} is conjectured to be OW-PCA secure under the Gap Diffie-Hellman assumption \cite{DBLP:conf/pkc/OkamotoP01}.

\subsubsection{Performance improvements}
The main advantage of our KEM construction over the Fujisaki-Okamoto transformation is the performance gains: our construction replaces re-encryption with computing a message authenticator, which is significantly faster. When instantiated with the underlying PKE routines of ML-KEM and the Poly1305 message authenticator, our construction $\text{ML-KEM}^+$ achieves on average 72\%-80\% reduction of CPU cycles needed for decapsulation while only incurring 2\%-7\% increase of CPU cycles for encapsulation when compared to ML-KEM.

\begin{table}[H]
    \centering
    \footnotesize

    \begin{tabular}{|p{1.2cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
        \hline
        & ML-KEM \newline 512 & $\text{ML-KEM}^+$ \newline 512 
        & ML-KEM \newline 768 & $\text{ML-KEM}^+$ \newline 768 
        & ML-KEM \newline 1024 & $\text{ML-KEM}^+$ \newline 1024 
        \\
        \hline
        Encap \newline (ccl/tick) 
        & 91467 & 93157 \newline (+1.8\%) 
        & 136405 & 146405 \newline (+7.3\%) 
        & 199185 & 205763 \newline (+3.3\%) 
        \\
        \hline
        Decap \newline (ccl/tick) 
        & 121185 & 33733 \newline (-72.2\%) 
        & 186445 & 43315 \newline (-76.8\%) 
        & 246245 & 51375 \newline (-79.1\%) 
        \\
        \hline
        CT size \newline (bytes) 
        & 768 & 784 \newline (+2.1\%) 
        & 1088 & 1104 \newline (+1.5\%) 
        & 1568 & 1584 \newline (+1.0\%) 
        \\
        \hline
    \end{tabular}

    \caption{$\text{ML-KEM}^+$ is instantiated with Poly1305}
\end{table}

We also implemented and measured the round trip time of key exchange protocols with various modes of authentication. When compared to ML-KEM, $\text{ML-KEM}^+$ achieves 24\%-28\% reduction of round trip time in unauthenticated key exchange (KE), 29\%-35\% reduction in unilaterally authenticated key exchange (UAKE), and 35\%-48\% reduction in mutually authenticated key exchange (AKE).

\begin{table}[H]
    \centering
    \footnotesize

    \begin{tabular}{|p{1.6cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
        \hline
        & ML-KEM \newline 512 & $\text{ML-KEM}^+$ \newline 512 
        & ML-KEM \newline 768 & $\text{ML-KEM}^+$ \newline 768 
        & ML-KEM \newline 1024 & $\text{ML-KEM}^+$ \newline 1024 
        \\
        \hline
        KE RTT \newline ($\us$) 
        & 92 & 70 \newline (-23.9\%) 
        & 135 & 99 \newline (-26.7\%) 
        & 193 & 138 \newline (-28.5\%) 
        \\
        \hline
        UAKE RTT \newline ($\us$) 
        & 145 & 103 \newline (-29.0\%) 
        & 215 & 144 \newline (-33.0\%) 
        & 310 & 202 \newline (-34.8\%) 
        \\
        \hline
        AKE RTT \newline ($\us$) 
        & 220 & 133 \newline (-39.5\%) 
        & 294 & 190 \newline (-35.4\%) 
        & 512 & 266 \newline (-48.0\%) 
        \\
        \hline
    \end{tabular}

    \caption{Key exchange round-trip times}
\end{table}

\subsection{Related works}
As mentioned at the beginning of section \ref{sec:introduction}, there is a large volume of previous proposals that combine cryptographic primitives with weak security properties into cryptographic primitives with strong security properties. Besides the Fujisaki-Okamoto transformation, another transformation that sees substantial adoption is \emph{Optimal Asymmetric Encryption Padding (OAEP)} \cite{bellare1995optimal}. Unfortunately, OAEP can only be applied to one-way trapdoor permutation, a rare occurrence. To this day, RSA remains the only viable candidate \cite{DBLP:conf/crypto/FujisakiOPS01}, though RSA-OAEP has been widely adopted after its standardization in PKCS\#1 v2.2 \cite{rfc8017}.

Securing an encryption scheme against chosen-ciphertext attack was a well-studied problem in the context of symmetric cryptography. Among the three generic patterns ``encrypt-then-MAC'', ``encrypt-and-MAC'', ``MAC-then-encrypt'', only the first one is formally analyzed and proved to be secure with any combination of symmetric cipher and MAC \cite{DBLP:conf/crypto/Krawczyk01}\cite{DBLP:conf/eurocrypt/CanettiK01}. ``MAC-then-encrypt'' is only secure with specific choices of symmetric cipher and MAC, and ``encrypt-and-MAC'' is generally not semantically secure since most MAC will leak information about the message it authenticates. Today's most well-established symmetric encryption schemes, including AES-GCM and ChaCha20-Poly1305, largely follow the pattern of ``encrypt-then-MAC''.

The idea of using a message authenticator to ensure ciphertext integrity was instantiated with ElGamal cryptosystem by Bellare and Rogaway in 1997 \cite{DBLP:conf/icics/BellareR97}, and its security was later reduced to the Gap Diffie-Hellman assumption by Abdulla, Bellare, and Rogaway in \cite{DBLP:conf/ctrsa/AbdallaBR01} in 2001. Our work can be seen as generalizing Bellare and Rogaway's scheme from ElGamal cryptosystem to all PKE with OW-PCA security.

@@

\subsection{Public-key encryption scheme}
\textbf{Syntax.} A public-key encryption scheme $\pke(\keygen, \encrypt, \decrypt)$ is a collection of three routines defined over some plaintext space $\mathcal{M}$ and some ciphertext space $\mathcal{C}$. $(\pk, \sk) \leftsample \keygen()$ is a randomized routine that returns a keypair. The encryption routine $\encrypt: (\pk, m) \mapsto c$ encrypts the input plaintext under the input public key. The decryption routine $\decrypt: (\sk, c) \mapsto m$ decrypts the input ciphertext under the input secret key. Where the encryption routine is randomized, we denote the randomness by $r \in \mathcal{R}$, where $\mathcal{R}$ is called the coin space. The decryption routine is assumed to always be deterministic. Some decryption routines can detect malformed ciphertext and output the rejection symbol $\bot$ accordingly.

\textbf{Correctness.} Following the definition in \cite{DBLP:conf/eurocrypt/DworkNR04} and \cite{hofheinz2017modular}, a $\pke$ is $\delta$-correct if:

@@

\textbf{Security.} We discuss the security of a $\pke$ using the sequence of games described in \cite{shoup2004sequences}. Specifically, we first define the $\texttt{OW-ATK}$ as they pertain to a public key encryption scheme. In later section we will define the $\texttt{IND-CCA}$ game as it pertains to a key encapsulation mechanism.
